---
title: "Density Estimation Demo"
output: html_notebook
---

```{r}
library(tidyverse)
giannis_shots <- read_csv("http://www.stat.cmu.edu/cmsac/sure/materials/data/xy_examples/giannis_2019_shots.csv")
head(giannis_shots)
```

```{r}
giannis_shots %>% 
  ggplot(aes(x = shot_distance)) +
  geom_histogram(binwidth = 1,
                 center = 0.5,
                 closed = "left") + 
  theme_bw()
```

Probability density function (PDF) is the derivative of the cumulative distribution function (CDF).

Histograms approximate the PDF with bins, and points are equally likely within a bin. (probability of a realization occuring within a bin follows a uniform distribution)

We want non-zero density between our observations, e.g. between 23 and 24 feet

## Kernel Density Estimation (KDE)

**Goal**: Estimate the PDF $f(x)$ for all possible values (assuming it is continuous / smooth)

Kernel density estimate: $f\hat(x) = \frac{1}{n}\displaystyle\sum^{n}_{i=1}\frac{1}{h}K_h(x-x_i)$

$n$: sample size, $x$: new point to esimate $f(x)$ (does NOT have to be in dataset!)

$h$: bandwidth, analogous to histogram bin width, ensures $f\hat(x)$ integrates to 1

$x_i$: ith observation in dataset

$K_h(x-x_i)$ is the Kernel function, creates weight given distance of ith observation from new points

as $|x - x_i| \rightarrow \infty$ then $K_h(x-x_i) \rightarrow 0$

Generally use the Gaussian kernel (normal distribution PDF) with variance = $h$

```{r}
giannis_shots %>% 
  ggplot(aes(x = shot_distance)) +
  geom_density() +
  geom_rug(alpha = 0.3) +
  theme_bw()
```

Use Gaussian reference rule (rule of thumb), $1.06\sigma n^{-\frac{1}{5}}$ where $\sigma$ is the observed standard deviation. This is the default bandwidth in R.

```{r}
# adjust multiplies the default bandwidth 
giannis_shots %>% 
  ggplot(aes(x = shot_distance)) +
  geom_density(adjust = 0.5) +
  geom_rug(alpha = 0.3) + theme_bw()
```

Use density curves and ECDF together!

```{r}
library(patchwork)

giannis_shot_dens <- giannis_shots %>% 
  ggplot(aes(x = shot_distance, color = is_shot_made)) +
  geom_density(adjust = 0.5) +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Density") +
  theme(legend.position = "none")

giannis_shot_ecdf <- giannis_shots %>% 
  ggplot(aes(x = shot_distance, color = is_shot_made)) +
  stat_ecdf() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Proportion of Giannis' shot attempts")

giannis_shot_dens / giannis_shot_ecdf + plot_layout(guides = "collect")
# stack on top: giannis_shot_dens / giannis_shot_ecdf
```

```{r}
library(ggridges)

giannis_shots %>% 
  ggplot(aes(x = shot_distance,
             y = shot_type)) +
  geom_density_ridges(rel_min_height = 0.01) +
  theme_bw()
```

```{r}
giannis_shots %>% 
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>% 
  filter(shot_y <= 30) %>% 
  ggplot(aes(x = shot_x, y = shot_y)) + 
  geom_point(alpha = 0.3) +
  geom_density_2d(adjust = 0.1) + 
  theme_bw() +
  coord_fixed()
```

```{r}
giannis_shots %>% 
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>% 
  filter(shot_y <= 30) %>% 
  ggplot(aes(x = shot_x, y = shot_y)) + 
  stat_density2d(h = 0.5, bins = 60,
                 aes(fill = after_stat(level)),
                 geom = "polygon") +
  scale_fill_gradient(low = "darkblue",
                      high = "darkorange") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  coord_fixed()
```

```{r}
library(hexbin)

giannis_shots %>% 
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>% 
  filter(shot_y <= 30) %>% 
  ggplot(aes(x = shot_x, y = shot_y, z = is_shot_made, group = -1)) + 
  stat_summary_hex(binwidth = c(2, 2),
                   color = "black",
                   fun = mean) +
  scale_fill_gradient(low = "darkblue",
                      high = "darkorange") +
  theme_bw() +
  theme(legend.position = "bottom") + 
  coord_fixed()
```