---
title: "EDA Demo: Sabermetrics"
output: html_notebook
---

```{r echo=FALSE}
library(tidyverse)
library(Lahman)
library(gt)
library(vcd)
Batting <- as_tibble(Batting)
```
A tibble is the tidyverse version of a data frame and contains additional features.

as_tibble() is preferred over as.tibble() (deprecated).

```{r}
dim(Batting)
```

```{r}
class(Batting)
```

```{r}
head(Batting)
```

```{r}
table(Batting$lgID)
```
## dplyr verbs
### filter
```{r}
mlb_batting <- filter(Batting, lgID %in% c("AL", "NL"))
# filter(Batting, (lgID == "AL") | (lgID == "NL"))
nrow(Batting) - nrow(mlb_batting)
```
### select
```{r}
sel_batting <- select(Batting, yearID, lgID, G, AB, H, BB, SO)
head(sel_batting, n = 3)
```
### arrange
```{r}
hr_batting <- arrange(Batting, desc(HR))
head(hr_batting)
```

### summarize
```{r}
summarize(Batting, max(stint), median(HR))
filter(Batting, stint == 5)
```

### mutate
```{r}
new_batting <- mutate(Batting, batting_avg = H / AB, so_to_bb = SO / BB)
select(new_batting, batting_avg, so_to_bb)
# this is annoying...
head(arrange(select(mutate(Batting, BA = H / AB), playerID, BA), desc(BA)), n = 3)
```

### pipe
```{r}
Batting %>% 
  filter(lgID %in% c("AL", "NL"),
         AB > 300) %>% 
  mutate(batting_avg = H / AB) %>% 
  arrange(desc(batting_avg)) %>% 
  select(playerID, yearID, batting_avg) %>% 
  slice(c(1, 2, 10, 100))
```
### group_by
```{r}
Batting %>% 
  filter(lgID %in% c("AL", "NL")) %>% 
  group_by(yearID) %>% 
  summarize(hr = sum(HR), so = sum(SO), bb = sum(BB)) %>% 
  arrange(desc(so)) %>% 
  slice(1:5)
```
```{r}
Batting %>% 
  filter(lgID %in% c("AL", "NL")) %>% 
  group_by(playerID) %>% 
  mutate(h_prop = H / sum(H, na.rm = TRUE)) %>% 
  arrange(desc(stint)) %>% 
  dplyr::select(playerID, stint, h_prop, H) %>% 
  ungroup()
```
### summarize_at
```{r}
year_batting_summary <- Batting %>% 
  filter(lgID %in% c("AL", "NL")) %>% 
  group_by(yearID) %>% 
  summarize_at(vars(H, HR, SO, BB, AB),
               sum, na.rm = TRUE) %>% 
  mutate(batting_avg = H / AB)
year_batting_summary
```

## More examples

When were the top 3 batting averages?
```{r}
year_batting_summary %>% 
  arrange(desc(batting_avg)) %>% 
  slice(1:3)
```

What is the strike out to walk ratio?
```{r}
year_batting_summary %>% 
  mutate(so_to_bb = SO / BB) %>% 
  arrange(so_to_bb) %>% 
  slice(c(1, n()))
```
What were the batting averages in 1876 and 2018?
```{r}
year_batting_summary %>% 
  select(yearID, batting_avg) %>% 
  rename(Year = yearID,
         `Batting AVG` = batting_avg) %>% 
  slice(c(1, n()))
```

```{r}
year_batting_summary %>% 
  select(yearID, batting_avg) %>% 
  rename(Year = yearID,
         `Batting AVG` = batting_avg) %>% 
  arrange(desc(`Batting AVG`)) %>% 
  slice(c(1:3, (n()-2):n())) %>% 
  gt() %>% 
  tab_header(
    title = "Best/worst MLB Seasons by AVG",
    subtitle = "Top/bottom three are presented"
  )
```
##ggplot2

```{r}
year_batting_summary %>% 
  ggplot(aes(x = yearID, y = HR)) +
  geom_point(aes(color = SO, size = BB)) +
  geom_line(color = "darkred",
            linetype = "dashed") +
  scale_color_gradient(low = "darkblue",
                       high = "darkorange") +
  labs(x = "Year", y = "Homeruns",
       color = "Strikeouts",
       size = "Walks",
       title = "The rise of MLB's three true outcomes",
       caption = "Data courtesy of Lahman") +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 15),
        axis.title = element_text(size = 10))
  # coord_cartesian() + (implicit)
  # scale_x_continuous() + (implicit)
  # scale_y_continuous() + (implicit)
  # scale_y_continuous(breaks = seq(0, 6000, by = 1000)) +
  # scale_x_reverse() +
  # scale_y_log10() +
  # geom_smooth() (smoothing regression)
```

### pivot the data

pivot_longer - gathers information spread out across variables, increase nrow() but decrease ncol()

pivot_wider - spreads information out from observations, decrease nrow() but increase ncol()

```{r}
year_batting_summary %>% 
  select(yearID, HR, SO, BB) %>% 
  rename(HRs = HR, Strikeouts = SO, Walks = BB) %>% 
  pivot_longer(HRs:Walks,
               names_to = "stat",
               values_to = "value") %>% 
  ggplot(aes(x = yearID, y = value)) +
  geom_line(color = "darkblue") +
  geom_point(color = "darkblue") +
  facet_wrap(~ stat, scales = "free_y", ncol = 1) +
  labs(x = "Year", y = "Total of statistic",
       title = "The rise of MLB's three true outcomes",
       caption = "Data courtesy of Lahman") +
  theme_bw() +
  theme(strip.background = element_blank())
```

### univariate data

#### 1D discrete

This is the empirical distribution of batted ball types from all hypothetical Mike Trout 2019 seasons.

Generate a point estimate of each batted ball type proportion and show the standard error. (Important to show uncertainty)

```{r}
trout_batted_balls <- read_csv("http://www.stat.cmu.edu/cmsac/sure/materials/data/xy_examples/trout_2019_batted_balls.csv")
head(trout_batted_balls)

# Frequency histogram
trout_batted_balls %>% 
  ggplot(aes(x = batted_ball_type)) +
  geom_bar(aes(y = ..count.. / sum(..count..))) +
  theme_bw()

# Relative frequency histogram
trout_batted_balls %>% 
  group_by(batted_ball_type) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(total = sum(count),
         prop = count / total,
         se = sqrt(prop * (1-prop) / total),
         lower = prop - 2 * se,
         upper = prop + 2 * se,
         batted_ball_type = fct_reorder(batted_ball_type, prop)) %>% 
  ggplot(aes(x = batted_ball_type)) +
  geom_bar(aes(y = prop),
           stat = "identity") +
  geom_errorbar(aes(ymin = lower,
                    ymax = upper),
                color = "red") +
  xlab("Batted Ball Type") +
  ylab("Proportion") +
  theme_bw()
```

However, we already know the population of Mike Trout's batted balls and a CI is not necessary, so use caution.

### 1D continuous

For continuous variables, we're interested in the shape, center, and spread of the distribution.

```{r}
# Boxplot
trout_batted_balls %>% 
  ggplot(aes(y = exit_velocity)) +
  geom_boxplot(aes(x = "")) +
  theme_bw() +
  coord_flip() +
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

Pros: displays outliers, percentiles, spread, skew
Cons: can't tell how many modes

```{r}
# Histogram
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity)) +
  geom_histogram(bins = 100) +
  theme_bw()
```

Still a summary.

```{r}
library(ggbeeswarm)
# Beeswarm plot (jittered dotplot)
trout_batted_balls %>% 
  ggplot(aes(y = exit_velocity)) +
  geom_beeswarm(aes(x = ""),
                cex = 3) +
  theme_bw() +
  coord_flip()
```

Shows all data points (good for small datasets).

```{r}
# Violin plot
trout_batted_balls %>% 
  ggplot(aes(x = "", y = exit_velocity)) +
  geom_violin() +
  geom_beeswarm(cex = 3) +
  geom_boxplot(width = 0.2,
               alpha = .6) + 
  theme_bw() +
  coord_flip()
```

Violin plots are nice because you can also see the distribution shape.

```{r}
# Empirical (sample) CDF
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity)) +
  stat_ecdf() +
  theme_bw()
```

Pros: ECDF displays all information in data (except for order), converges in distribution, easy to interpret
Cons: Not very popular

```{r}
# Rug plot + histogram
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity)) +
  geom_histogram() +
  geom_rug(alpha = 0.3) +
  theme_bw()

# Rug + scatterplot
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity,
             y = launch_angle)) +
  geom_point() +
  geom_rug(alpha = 0.3) +
  theme_bw()
```

Useful when added to other visualizations, e.g. ECDF

### bivariate data

#### 2D categorical

```{r}
trout_batted_balls <- trout_batted_balls %>% 
  filter(pitch_type != "null") %>% 
  mutate(pitch_type = fct_recode(pitch_type,
                                 "Changeup" = "CH",
                                 "Breaking ball" = "CU",
                                 "Fastball" = "FF",
                                 "Fastball" = "FC",
                                 "Fastball" = "FS",
                                 "Fastball" = "FT",
                                 "Breaking ball" = "KC",
                                 "Fastball" = "SI",
                                 "Breaking ball" = "SL"
                                 ))
```

```{r}
trout_batted_balls %>% 
  ggplot(aes(x = batted_ball_type,
             fill = pitch_type)) +
  geom_bar(position = "dodge") +
  theme_bw()
```

Stacked bar charts emphasize the marginal distribution.

Side-by-side bar charts emphasize the conditional distribution.

```{r}
# contingency table - frequency
trout_batted_balls %>% 
  group_by(batted_ball_type, pitch_type) %>% 
  summarize(joint_freq = n()) %>% 
  pivot_wider(names_from = batted_ball_type,
              value_from = joint_freq)

# contingency table - probability
trout_batted_balls %>% 
  group_by(batted_ball_type, pitch_type) %>% 
  summarize(joint_prob = n() / nrow(trout_batted_balls)) %>% 
  pivot_wider(names_from = batted_ball_type,
              values_from = joint_prob) %>% 
  gt()

```

Marginal distribution - row/column sums
Conditional distribution - individual cell/marginal 
Two variables X, Y are independent iff P(x)P(y) = P(x, y).

```{r}
# visualizing independence
mosaic(~ pitch_type + batted_ball_type,
       data = trout_batted_balls)
```

Spine chart of spine charts
height = marginal distribution of pitch_type
width = conditional distribution of batted_ball_type | pitch_type
area = joint distribution

```{r}
# side-by-side violins
trout_batted_balls %>% 
  ggplot(aes(x = pitch_type,
             y = exit_velocity)) +
  geom_violin() +
  geom_boxplot(width = .4) +
  theme_bw()
```

```{r}
# ecdf
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity,
             color = pitch_type)) +
  stat_ecdf() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
# stacked histogram
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity,
             fill = pitch_type)) +
  geom_histogram() + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

```{r}
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity,
             color = pitch_type)) +
  geom_freqpoly() +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity)) + 
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ pitch_type, ncol = 2)
```

```{r}
trout_batted_balls %>% 
  ggplot(aes(x = exit_velocity)) + 
  geom_histogram() +
  theme_bw() +
  facet_grid(pitch_type ~., margins = TRUE)
```

### multivariable data
```{r}
trout_batted_balls %>% 
  ggplot(aes(x = pitch_type,
             fill = batted_ball_type)) +
  geom_bar() +
  theme_bw() +
  facet_wrap(~ outcome, ncol = 5) + 
  theme(legend.position = "bottom")
```