---
title: "MLB Regression Analysis"
author: "Thea Sukianto, James Hyman"
date: "6/19/2020"
output: pdf_document
---

```{r}
library(tidyverse)
library(patchwork)
library(GGally)
library(car)
library(ggcorrplot)
```

## Exploratory Data Analysis

```{r message = FALSE}
baseball_batting <- read_csv("http://www.stat.cmu.edu/cmsac/sure/materials/data/regression_projects/baseball_batting.csv") %>% 
  dplyr::rename(Index = X1) %>% 
  dplyr::mutate(WPA_net = WPA_plus + WPA_minus,
                Name = as.factor(Name),
                Team = as.factor(Team),
                playerid = as.factor(playerid),
                Season = as.factor(Season),
                strike_percent = Strikes/Pitches)

str(baseball_batting)
```

Each row of `baseball_batting` represents season MLB batting statistics for an individual player between 2000-2019. There are a total of 1,429 observations and 39 variables. 

Of these 39 variables, four are factors that represent a player's `Name`, `Team`, `playerid`, and `Season`. There are missing values in `Team` that may signify player trading.

The remaining variables are numeric and represent various season-level batting statistics. These can be separated into two categories, "in-game" statistics and "calculated" statistics.

For example, one of the statistics in the "calculated" category is `WPA` (win probability added). We add an additional column, `WPA_net`, as the sum of `WPA_plus` and `WPA_minus` to show the net change in win probability per player per season.


### Exploring WPA

```{r}
# View the overall distribution of Net WPA
baseball_batting %>% 
  ggplot(aes(WPA_net)) +
  geom_histogram() +
  labs(x = "Net Win Probability Added (%)",
       y = "Count",
       title = "Distribution of MLB Net Win Probability Per Player",
       subtitle = "Most players balanced between positive and negative WPA") +
  theme_bw()
```

The distribution of net win probability added is unimodal and roughly symmetric (centered around 0), with a slight skew to the right. This indicates that the average player has a balance between positive and negative win probability added, but there are a few all-stars with a net win probability added above 5. It appears that net win probability would be a good candidate for the response variable.


#### Potential Contributors to Net Win Probability Added

```{r}
bb_net_vars <- baseball_batting %>% 
  dplyr::select(WPA_net, H, HR, R, RBI, AVG, BB, SLG, OBP, SO, SB, CS, strike_percent) %>% 
  dplyr::rename(`Net WPA` = WPA_net,
                `Strike %` = strike_percent)

ggcorrplot(round(cor(bb_net_vars), 2), type = "lower", lab = TRUE, lab_size = 3,
           title = "Selected Batting Statistics vs. Net WPA Correlation Matrix") +
  labs(subtitle = "SLG and OBP have strongest linear relationship with Net WPA")
```

Based on the correlation matrix above, we make a conservative decision to choose variables $-.6 < r < 6$ as to avoid collinearity. Therefore, when defining linear models, (RBI, HR), (AVG, H), (SLG, HR), (RBI, SLG), (R, SLG), (SLG, OBP), (BB, OBP), (AVG, OBP), (Strike %, BB), (Strike %, OBP), (H, R) will not be put in the same model. We leave SO, SB, and CS out of potential models since the correlation values are so low.

```{r}
# View scatterplots for "in-game" statistics
ggpairs(baseball_batting, columns = c("WPA_net", "H", "HR", "R", "RBI", "BB"), mapping = aes(alpha = 0.3), columnLabels = c("Net WPA", "H", "HR", "R", "RBI", "BB")) + labs(title = "Relationships Between Selected In-Game Statistics and Net WPA", subtitle = "R and RBI have strongest linear relationship with Net WPA")
```


```{r}
# Plot "calculated" statistics

ggpairs(baseball_batting, columns = c("WPA_net", "AVG", "SLG", "OBP", "strike_percent"), mapping = aes(alpha = 0.3), columnLabels = c("Net WPA", "AVG", "SLG", "OBP", "Strike %")) + labs(title = "Relationships Between Selected Aggregate Statistics and Net WPA", subtitle = "SLG and OBP have strongest linear relationship with Net WPA")
```

## Variable Selection and Model Selection

```{r}
# Modify cross-validation code from lecture to do leave-one-season-out cross validation
get_cv_preds <- function(model_formula, data = baseball_batting) {
  # generate holdout predictions for every row based season
  purrr::map_dfr(unique(data$Season), 
          function(holdout) {
            # Separate test and training data:
            test_data <- data %>%
              filter(Season == holdout)
            train_data <- data %>%
              filter(Season != holdout)
            # Train model:
            reg_model <- lm(as.formula(model_formula), data = train_data)
            # Return tibble of holdout results:
            tibble(test_preds = predict(reg_model, newdata = test_data),
                   test_actual = test_data$WPA_net,
                   Season = holdout) 
          })
}
```


```{r}
# Select variables for potential models based on EDA

# Models with "in-game" statistics
h_hr_bb <- get_cv_preds("WPA_net ~ H + HR + BB")
h_rbi_bb <- get_cv_preds("WPA_net ~ H + RBI + BB")
r_hr_bb <- get_cv_preds("WPA_net ~ R + HR + BB")
r_rbi_bb <- get_cv_preds("WPA_net ~ R + RBI + BB") 

# Models with "calculated" statistics
avg_slg_strikes <- get_cv_preds("WPA_net ~ AVG + SLG + strike_percent")

# Combination - 3 variables
h_hr_obp <- get_cv_preds("WPA_net ~ H + HR + OBP")
h_hr_strikes <- get_cv_preds("WPA_net ~ H + HR + strike_percent")
h_rbi_obp <- get_cv_preds("WPA_net ~ H + RBI + OBP + strike_percent")
h_rbi_strikes <- get_cv_preds("WPA_net ~ H + HR + strike_percent")
h_bb_slg <- get_cv_preds("WPA_net ~ H + BB + SLG")
r_rbi_obp <- get_cv_preds("WPA_net ~ R + RBI + OBP")
rbi_bb_avg <- get_cv_preds("WPA_net ~ RBI + BB + AVG")
rbi_bb_strikes <- get_cv_preds("WPA_net ~ RBI + BB + strike_percent")

# Combination models - 4 variables
r_rbi_avg_strikes <- get_cv_preds("WPA_net ~ R + RBI + AVG + strike_percent")
r_hr_avg_strikes <- get_cv_preds("WPA_net ~ R + RBI + AVG + strike_percent")
```


```{r}
bind_rows(mutate(h_hr_bb, type = "H + HR + BB"),
          mutate(h_rbi_bb, type = "H + RBI + BB"),
          mutate(r_hr_bb, type = "R + HR + BB"),
          mutate(r_rbi_bb, type = "R + RBI + BB"),
          mutate(avg_slg_strikes, type = "AVG + SLG + Strike %"),
          mutate(h_hr_obp, type = "H + HR + OBP"),
          mutate(h_hr_strikes, type = "H + HR + Strike %"),
          mutate(h_rbi_obp, type = "H + RBI + OBP"),
          mutate(h_rbi_strikes, type = "H + RBI + Strike %"),
          mutate(h_bb_slg, type = "H + BB + SLG"),
          mutate(r_rbi_obp, type = "R + RBI + OBP"),
          mutate(rbi_bb_avg, type = "RBI + BB + AVG"),
          mutate(rbi_bb_strikes, type = "RBI + BB + Strike %"),
          mutate(r_rbi_avg_strikes, type = "R + RBI + AVG + Strike %"),
          mutate(r_hr_avg_strikes, type = "R + HR + AVG + Strike %")) %>%
  group_by(type) %>%
  summarize(rmse = sqrt(mean((test_actual - test_preds)^2))) %>%
  mutate(type = fct_reorder(type, rmse)) %>%
  ggplot(aes(x = type, y = rmse)) +
  labs(x = "Model",
       y = "Root Mean Squared Error (RMSE)",
       title = "Leave-one-season-out Cross Validation Model Performance",
       subtitle = "Models considering both RBI and OBP performed best") + 
  coord_flip() + 
  geom_point() + 
  theme_bw()
```

## Model Evaluation

```{r}
# Create a linear model from the best CV RMSE and view diagnostic plots
r_rbi_obp_lm <- lm("WPA_net ~ R + RBI + OBP", data = baseball_batting)
plot(r_rbi_obp_lm)
```

```{r}
# View summary of best model
summary(r_rbi_obp_lm)
```
```